Local Spectral Clustering of Density Upper Level Sets

### 2019-05-23 neurips/local_spectral_density_clustering_neurips.tex
Temp commit

To motivate these conditions, \citet{zhu2013} show for arbitrary graph $G =
(V,E)$ and subset of vertices $S \subseteq V$, the PPR estimate \smash{$\Cest$}
of subset $S$ satisfies, for a constant $c>0$,
\begin{equation}
\label{eqn: graph_symmetric_set_difference_1}
\vol(\Cest \setminus S; G) + \vol(S \setminus \Cest; G) \leq c \bigl( \Phi(S,G)
\cdot \tau_{\infty}(G[S])\bigr) \vol(S;G),
\end{equation}
where $\Phi(S;G)$ is the normalized cut of $S$ (as defined in \eqref{eqn:
  normalized_cut}), and \smash{$\tau_{\infty}(G[S])$} is called the \emph{mixing
  time} of a random walk over the induced subgraph $G[S]$ (to be defined
precisely later, in \eqref{eqn: mixing_time}).  The left-hand side in
\eqref{eqn:  graph_symmetric_set_difference_1} resembles a (degree-weighted) 
form of the symmetric set difference metric in \eqref{eqn:
  misclassification_rate}.  As we will show in Section \ref{sec:
  analysis}, the conditions \ref{asmp: bounded_density}-\ref{asmp:
  bounded_volume} allow us to upper bound the normalized cut 
\smash{$\Phi(\Csig[\Xbf]; G_{n,r})$}, and the mixing time 
\smash{$\tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$}.  (Informally, 
\ref{asmp: cluster_separation}, \ref{asmp: low_noise_density} yield an upper
bound on \smash{$\cut(\Csig[\Xbf]; G_{n,r})$}, and \ref{asmp: bounded_density}
yields a lower bound on \smash{$\vol(\Csig[\Xbf]; G_{n,r})$}; together with 
\ref{asmp: bounded_volume}, this gives an upper bound on the normalized cut.  On 
the other hand, \ref{asmp: bounded_density}, \ref{asmp: embedding} preclude
bottlenecks in the induced subgraph \smash{$G_{n,r}[\Csig[\Xbf]]$}, and combined
with \ref{asmp: embedding}, this leads to an upper bound on the mixing time over 
this subgraph.)

@@

The primary technical contribution of our work is showing that the geometric conditions \ref{asmp: bounded_density} - \ref{asmp: bounded_volume} translate to meaningful bounds on the normalized cut and mixing time of $\Csig[\Xbf]$ in $G_{n,r}$. In doing so, we elaborate on how some of the geometric conditions introduced in Section \ref{sec: consistent_cluster_estimation_with_ppr} contribute to the difficulty of the clustering problem.

\paragraph{Normalized cut} We start with a finite sample upper bound on the normalized cut \eqref{eqn: normalized_cut} of 
$\Cset_\sigma[\Xbf]$. For simplicity, we write $\Phi_{n,r}(\Csig[\Xbf]) := \Phi(\Csig[\Xbf]; G_{n,r})$.

\begin{theorem}
	\label{thm: conductance_upper_bound}
	Fix $\lambda > 0$, and let $\Cset \in \Cbb_f(\lambda)$ satisfy
	Assumptions \ref{asmp: bounded_density}-\ref{asmp: low_noise_density}, and \ref{asmp: bounded_volume} for some 
	$r, \sigma, \lambda_{\sigma}, c_0, \gamma > 0$ (no bound on maximum density is needed). 
	Then for any $0 < \delta < 1$, $\epsilon > 0$, if
	\begin{equation}
	\label{eqn: conductance_sample_complexity}
	n \geq \frac{(2+\epsilon)^2\log(3/\delta)}{\epsilon^2}\left(\frac{25}
	{6 \lambda_{\sigma}^2\nu(\Csig) \nu_d r^d}\right)^2,
	\end{equation}
	then
	\begin{equation}
	\label{eqn: conductance_additive_error_bound}
	\frac{\Phi_{n,r}(\Csig[\Xbf])}{r} \leq c_1 \frac{d}{\sigma}
	\frac{\lambda}{\lambda_{\sigma}} \frac{(\lambda_{\sigma} -
		c_0\frac{r^{\gamma}}{\gamma+1})}{\lambda_{\sigma}} + \epsilon, 
	\end{equation}
	with probability at least $1-\delta$ (where $c_1 > 0$ is a universal constant).
\end{theorem}

\begin{remark}
	\label{rmk: diameter}
	Observe that the diameter $\rho$ is absent from Theorem \ref{thm: conductance_upper_bound}, in contrast to the difficulty function $\kappa(\Cset)$, which worsens (increases) as $\rho$ increases. This phenomenon reflects established wisdom regarding spectral partitioning algorithms more generally \cite{guattery1995, hein2010}, albeit newly applied to the density clustering setting. It suggests that if the diameter $\rho$ is large, \pprspace may fail to recover $\Csig[\Xbf]$ even when $\Cset$ is sufficiently well-conditioned to ensure $\Csig[\Xbf]$ has a small normalized cut in $G_{n,r}$. This intuition will be supported by simulations in Section \ref{sec: experiments}.
\end{remark}

\paragraph{Mixing time.}
For $S \subseteq V$, denote by $G[S] = (S, E_S)$ the subgraph induced by 
$S$ (where the edges are $E_S = E \cap (S \times S)$). Let $\Wbf_S$ be the (lazy) random walk matrix over $G[S]$, and write 
$$
q_{v}^{(t)}(u) = e_v\Wbf_S^t e_u
$$
for the $t$-step transition probability of the lazy random walk over $G[S]$
originating at $v \in V$. Also write \smash{$\pi = (\pi(u))_{u \in S}$}
for the stationary distribution of this random walk.  (As
$\Wbf_S$ is the transition matrix of a lazy random walk, it is well-known that a unique stationary distribution exists and is given by
\smash{$\pi(u) = (\Dbf_S)_{uu}/\vol(S; G[S])$}, where $\Dbf_S$ is the degree matrix of $G[S]$.)

Then, the \emph{relative pointwise mixing time} of $G[S]$ is 
\begin{equation}
\label{eqn: mixing_time}
\tau_{\infty}(G[S]) = \min\set{ t: \frac{\pi(u) - q_{v}^{(t)}(u)
	}{\pi(u)} \leq \frac{1}{4}, 
	\; \text{for $u,v \in V$}}. 
\end{equation}
In the following theorem, we give an asymptotic (in the number of vertices $n$) upper bound on $\tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$.

>>>>

To motivate these conditions, \citet{zhu2013} show for arbitrary graph $G =
(V,E)$ and subset of vertices $S \subseteq V$, the PPR estimate \smash{$\Cest$}
of subset $S$ satisfies, for a constant $c>0$,
\begin{equation}
\label{eqn: graph_symmetric_set_difference_1}
\vol(\Cest \setminus S; G) + \vol(S \setminus \Cest; G) \leq c \bigl( \Phi(S,G)
\cdot \tau_{\infty}(G[S])\bigr) \vol(S;G),
\end{equation}
where $\Phi(S;G)$ is the normalized cut of $S$ (as defined in \eqref{eqn:
  normalized_cut}), and \smash{$\tau_{\infty}(G[S])$} is called the \emph{mixing
  time} of a random walk over the induced subgraph $G[S]$ (to be defined
precisely later, in \eqref{eqn: mixing_time}).  The left-hand side in
\eqref{eqn:  graph_symmetric_set_difference_1} resembles a (degree-weighted) 
form of the symmetric set difference metric in \eqref{eqn:
  misclassification_rate}.  As we will show in Section \ref{sec:
  analysis}, the conditions \ref{asmp: bounded_density}--\ref{asmp:
  bounded_volume} allow us to upper bound the normalized cut 
\smash{$\Phi(\Csig[\Xbf]; G_{n,r})$}, and the mixing time 
\smash{$\tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$}.  (Informally, 
\ref{asmp: cluster_separation}, \ref{asmp: low_noise_density} yield an upper
bound on \smash{$\cut(\Csig[\Xbf]; G_{n,r})$}, and \ref{asmp: bounded_density}
yields a lower bound on \smash{$\vol(\Csig[\Xbf]; G_{n,r})$}; together with 
\ref{asmp: bounded_volume}, this gives an upper bound on the normalized cut.  On 
the other hand, \ref{asmp: bounded_density}, \ref{asmp: embedding} preclude
bottlenecks in the induced subgraph \smash{$G_{n,r}[\Csig[\Xbf]]$}, and combined
with \ref{asmp: embedding}, this leads to an upper bound on the mixing time over 
this subgraph.)

@@

The primary technical contribution of our work is showing that the geometric
conditions \ref{asmp: bounded_density}--\ref{asmp: bounded_volume} translate to 
meaningful bounds on the normalized cut and mixing time of $\Csig[\Xbf]$ in
$G_{n,r}$. In doing so, we elaborate on how some of the geometric conditions
introduced in Section \ref{sec: consistent_cluster_estimation_with_ppr}
contribute to the difficulty of the clustering problem.

\paragraph{Normalized cut.} We start with a finite sample upper bound on the
normalized cut \eqref{eqn: normalized_cut} of $\Cset_\sigma[\Xbf]$. For
simplicity, we write $\Phi_{n,r}(\Csig[\Xbf]) := \Phi(\Csig[\Xbf]; G_{n,r})$.

\begin{theorem}
  \label{thm: conductance_upper_bound}
  Fix $\lambda > 0$, and assume $\Cset \in \Cbb_f(\lambda)$ satisfies
  Assumptions \ref{asmp: bounded_density}--\ref{asmp: low_noise_density}, 
  \ref{asmp: bounded_volume} for some $r, \sigma, \lambda_{\sigma}, c_0, \gamma
  > 0$ (no bound on maximum density is needed). Then for any $0 < \delta < 1$,
  $\epsilon > 0$, if 
  \begin{equation}
    \label{eqn: conductance_sample_complexity}
    n \geq \frac{(2+\epsilon)^2\log(3/\delta)}{\epsilon^2}\left(\frac{25}
      {6 \lambda_{\sigma}^2\nu(\Csig) \nu_d r^d}\right)^2,
  \end{equation}
  then
  \begin{equation}
    \label{eqn: conductance_additive_error_bound}
    \frac{\Phi_{n,r}(\Csig[\Xbf])}{r} \leq c_1 \frac{d}{\sigma}
    \frac{\lambda}{\lambda_{\sigma}} \frac{(\lambda_{\sigma} -
      c_0\frac{r^{\gamma}}{\gamma+1})}{\lambda_{\sigma}} + \epsilon, 
  \end{equation}
  with probability at least $1-\delta$ (where $c_1 > 0$ is a universal constant).
\end{theorem}

\begin{remark}
  \label{rmk: diameter}
  Observe that the diameter $\rho$ is absent from Theorem \ref{thm:
    conductance_upper_bound}, in contrast to the condition number
  $\kappa(\Cset)$, which worsens (increases) as $\rho$ increases. This
  reflects established wisdom regarding spectral partitioning
  algorithms more generally \citep{guattery1995, hein2010}, albeit newly applied
  to the density clustering setting. It suggests that if the diameter $\rho$ is
  large, PPR may fail to recover $\Csig[\Xbf]$ even when $\Cset$ is
  sufficiently well-conditioned to ensure $\Csig[\Xbf]$ has a small normalized
  cut in $G_{n,r}$. This intuition will be supported by simulations in Section
  \ref{sec: experiments}. 
\end{remark}

\paragraph{Mixing time.} For $S \subseteq V$, denote by $G[S] = (S, E_S)$ the
subgraph induced by $S$ (where the edges are $E_S = E \cap (S \times S)$). Let
$\Wbf_S$ be the (lazy) random walk matrix over $G[S]$, and write  
$$
q_{v}^{(t)}(u) = e_v\Wbf_S^t e_u
$$
for the $t$-step transition probability of the lazy random walk over $G[S]$
originating at $v \in V$. Also write \smash{$\pi = (\pi(u))_{u \in S}$}
for the stationary distribution of this random walk.  (As
$\Wbf_S$ is the transition matrix of a lazy random walk, 
it is well-known that a unique stationary distribution exists and is given by 
\smash{$\pi(u) = (\Dbf_S)_{uu}/\vol(S; G[S])$}, where we write $\Dbf_S$ for the  
degree matrix of $G[S]$.) We define the \emph{mixing time} of $G[S]$ as
\begin{equation}
\label{eqn: mixing_time}
\tau_{\infty}(G[S]) = \min\set{ t: \frac{\pi(u) - q_{v}^{(t)}(u)}
  {\pi(u)} \leq \frac{1}{4}, \; \text{for $u,v \in V$}}. 
\end{equation}
Next, we give an asymptotic (in the number of vertices $n$) upper bound on
$\tau_{\infty}(G_{n,r}[\Csig[\Xbf]])$.
